---
title: "Exploratory DataFest Analysis"
author: '@jimmylovestea'
date: "UCLA DataFest Workshop II"
output: 
  ioslides_presentation: 
    transition: faster
editor_options: 
  chunk_output_type: console
---

<!-- Custom JS/CSS to add the fixed footer -->
<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.12.2/jquery.min.js"></script>

<script>
    $(document).ready(function() {
    $('slide:not(.title-slide, .backdrop, .segue)').append('<footer></footer>');    
    $('footer').attr('label', 'https://github.com/jimmylovestea/DataFest2017');

  })
</script>

<style>
  footer:after {
    font-size: 12pt;
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 20px;
    line-height: 1.9;
    display: block;
  }
</style>
<!-- End custom JS/CSS for fixed footer -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workshop Focus

- Practice exploring data using the `tidyverse` set of packages.
    - Manipulate/sort data with `dplyr`.
    - Convert wide data to long data with `tidyr`.
    - Make plots with `ggplot2`.
    - String commands together with pipes (`%>%`)

## From last time:

_Workshop I_ focused on meeting with your team to start getting organized.

- How will the team share files? 
    - I suggested Dropbox.
- How will the team spend the first part of night one? 
    - One person reads the data and saves it. 
    - Other team members read any documentation connected to the data.
    - Everyone starts brain-storming questions to try and answer with the data.

## From last time:

This workshop focuses on what to do once you:

- Have read in (and saved as an .rda) the raw data.
- Have spent some time cleaning the variable names, checking the `class()` of the variables, etc.
- Understand what the individual variables represent in the context of the data.

Specifically, exploring your data with the `tidyverse`.

## `tidyverse`?

The `tidyverse` is alternative approach to coding in **R**.

- Code is still written in **R**.
- Code is still executed in **R**.
- The style of how the code is written is slightly different.

Specifically:

- Code utilizes pipes to take the output of one function and place it into a second function.
- Focus is on manipulating data frames.
- Packages from the `tidyverse` tend to work really well with each other.

## `tidyverse`?

### Example: Calculate the mean of `y` conditioned on the value of `x`.

```{r, echo = TRUE}
df <- data.frame(x = c("A", "A", "B", "B", "A", "C"),
                 y = c(1, 8, 6, 9, 7, 3))
df
```

## `tidyverse`?

### `base` **R**

```{r}
mean(df$y[df$x == "A"])
mean(df$y[df$x == "B"])
mean(df$y[df$x == "C"])
```

## `tidyverse`?

### `tidyverse`

```{r, warning = FALSE, message = FALSE}
library(dplyr)
df %>% 
  group_by(x) %>% 
  summarize(means = mean(y))
```


# Exploring your data.

## Goals

When doing EDA, you want to:

- Learn about the distributions of variables.
- Look for how variables are related.
- Use what you learn to generate questions about the data...
- ... and then use more EDA to look for answers.

## Fuel Economy Data

Our data deals with fuel economy in vehicles from 1984-present.

- https://www.fueleconomy.gov/feg/ws/index.shtml#vehicle
    - Look for: **Download the Vehicle Data**
    - Variable definitions are under the **Data Description** label.
    - Data are also on Github.

## {.flexbox .vcenter}

_Follow along as we read in our data, look through the vehicle definitions and make plots with `ggplot2`_

## Summary

We've just demo'd:

- Using the `readr` package to read in the data.
    - Also, how to solve variable encoding issues from within `readr` functions.
- Looked through the variable definitions and generated a question.
- Used plots made with `ggplot2` to answer our question.

## `ggplot2`

`ggplot2` focuses on building plots in layers.

- Start with data.
- Include some aesthetics
    - _What pieces of the data will be plotted?_
- Add a `geometry`
    - _Where should each aesthetic be placed in the plot?_
- Iterate:
    - Change scales/axes.
    - Include additional layers/geoms.
    - Change theme elements.

## On Your Own

Accomplish the following:

1. Go to http://ggplot2.tidyverse.org/reference/
    - Click on a `geom_`, scroll to the examples.
2. Pick a variable from the data that you can use with the `geom_` you chose.
3. Use `ggplot2`, the `geom_` and the variable to create a plot.
    - What does the plot teach you about that variable?
    - Based on this plot, what else might you want to learn from the data?

## `tidyr`

When doing EDA, we often want to compare distributions of different variables.

- This can be a hassle to do with just `ggplot2` because `ggplot2` likes to use _long_ and not _wide_ data.
- `tidyr` makes converting wide to long to wide data easy.

## Wide data

```{r, echo = FALSE}
df <- data.frame(name = c("Luke", "Leia", "Han", "Chewie"),
                 primary = c("Reading", "Reading", "Looking in Mirror", "Roaring"),
                 secondary = c("Brooding", "Taking down Empires", "Smuggling", "Smuggling"))
df
```

## Long data

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyr)
gather(df, key = preference, like, c(primary, secondary))
```


## {.flexbox .vcenter}

_Follow along as we convert our wide data to long and then use the long data to facet in ggplot_

## Summary

We've just demo'd:

- Using `tidyr` to create a long data set.
- How to use pipe (`%>%`) to push the data into `ggplot2`
- Create a faceted histogram of our data.

## Pipes (`%>%`)

Pipes are one of the primary benefits of the `tidyverse`.

- Pipes take the output from one function (on the left).
    - And then pipes the output to the first argument of a new function (on the right).
    
### this_output() %>% will_go(right_here)

## `dplyr`

`dplyr` makes doing data manipulations easy.

- You can `select()` or `arrange()` your data.
- You can `mutate()` or `summarize()` your data.
- You can also `filter()` your data.

## {.flexbox .vcenter}

_Follow along as we combine the data manipulations from `dplyr` with `ggplot2`_

## Grouped operations

Another big benefit of `dplyr` is its ability to perform group operations.

- Group the data by a categorical data.
- Summarize each group.

## {.flexbox .vcenter}

_Follow along as we perform some group operations to calculate the average and standard deviations of MPG per year_

## Other `dplyr` features

`dplyr` contains another set of significant features (which I didn't think I'd have time to present on today). These features are about joining data together.

- Highly recommend reading & following along with the examples of joining data with `dplyr` here:
    - http://r4ds.had.co.nz/relational-data.html

## On Your Own

Accomplish the following:

1. Compare the distributions of MPG (`highway08`) for Cadillacs and Toyotas in a single, faceted plot.
2. Overlay the density of MPG (`highway08`) for Cadillacs and Toyotas in a single_non-faceted plot.
3. Plot the average MPG (by year) as a line. Include the the points in the plot.

# Overview

## 

The lessons to take away from these first two workshops:

- Start prepping for DataFest before the data is revealed.
- Have a plan for night one:
    - Learn about data.
    - Read it into **R** and share it as an `.rda`.
    - Generate ideas for your project by exploring the data.
    - Stop. And walk away. Talk about how you want to proceed. Establish Goals. Divide tasks up to accomplish the goals. Keep everyone involved.


## {.flexbox .vcenter}

### Questions?

## {.flexbox .vcenter}

### <3 Thanks for attending! <3